---
title: "BDSI - Data Manipulation with dplyr and nycflights13"
author: "Jacob Gladfelter"
editor: visual
format: 
  html:
    theme: darkly
---

## Overview

This document demonstrates data manipulation using `dplyr` and visualization with `ggplot2` on the `nycflights13` data set.

## Install required packages

```{r results="hide", message=FALSE, warning=FALSE}
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
if (!requireNamespace("nycflights13", quietly = TRUE)) {
  install.packages("nycflights13")
}
library(dplyr)
library(tidyverse)
library(nycflights13)
library(ggplot2)
library(ggdark)
```

## Preview the flights data

```{r warning=FALSE}
head(flights)
glimpse(flights)
```

## Filtering rows: Flights on January 1st

```{r warning=FALSE}
jan1_flights <- flights %>% filter(month == 1, day == 1)
head(jan1_flights)
```

## Selecting columns: Only flight and delay info

```{r warning=FALSE}
delay_info <- flights %>% select(year, month, day, dep_delay, arr_delay)
head(delay_info)
```

## Mutating columns: Calculate total delay

```{r warning=FALSE}
flights_delay <- flights %>% mutate(total_delay = dep_delay + arr_delay)
head(flights_delay)
```

## Summarizing: Average delay by airline

```{r warning=FALSE}
avg_delay <- flights %>% group_by(carrier) %>% summarise(mean_arr_delay = mean(arr_delay, na.rm = TRUE))
avg_delay
```

## Plot: Average arrival delay by airline

```{r warning=FALSE}
ggplot2::ggplot(avg_delay, aes(x = carrier, y = mean_arr_delay)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Arrival Delay by Airline", x = "Carrier", y = "Mean Arrival Delay (min)") +
  dark_theme_gray()
```

## Plot: Distribution of departure delays

```{r warning=FALSE}
ggplot2::ggplot(flights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 10, fill = "orange", color = "black") +
  xlim(-50, 300) +
  labs(title = "Distribution of Departure Delays", x = "Departure Delay (min)", y = "Count") +
  dark_theme_gray()
```

## Plot: Relationship between departure and arrival delay

```{r warning=FALSE}
ggplot2::ggplot(flights, aes(x = dep_delay, y = arr_delay)) +
  geom_point(alpha = 0.2) +
  xlim(-50, 300) + ylim(-50, 300) +
  labs(title = "Departure vs Arrival Delay", x = "Departure Delay (min)", y = "Arrival Delay (min)") +
  dark_theme_gray()
```
